Accelerated oblique random survival forests

### 2022-07-09 paper/jmlr/main.tex
A.3 shows values from VI

Risk prediction can reduce the burden of disease by educating patients and providers and guiding strategies to prevent and treat disease in a wide range of medical domains \citep{moons2012riskII, moons2012riskI}. The random survival forest (RSF), a supervised learning algorithm that can engage with censored outcomes, is frequently used for risk prediction. Notable characteristics of the RSF include uniform convergence of its ensemble survival function to the true population survival function when the predictor space is discrete \citep{ishwaran2010consistency}. In addition, software implementing the RSF is freely available, extremely efficient, and full of tools to interpret and explain the RSF \citep{randomForestSRC, ranger, hothorn2010party}. However, there remains considerable potential to improve the RSF in risk prediction tasks where training samples are not large enough to guarantee asymptotic properties or predictor spaces are non-discrete (\ie, predictors are continuous).

RSFs may be axis based or oblique. The axis based RSF uses a single predictor whereas the oblique RSF uses a linear combination of predictors to create branches in trees. While axis based decision boundaries are always perpendicular to the axis of the relevant predictor, linear combinations of predictors create oblique decision boundaries that are neither parallel nor perpendicular to axes of their contributing predictors. Prior work has found the oblique RSF has higher prediction accuracy than the axis based RSF in general benchmarks \citep{jaeger2019oblique} and that oblique splitting is particularly effective when predictors are continuous \citep{menze2011oblique}. However, existing methods to implement oblique splitting typically use fully trained models in each non-leaf node to identify linear combinations of predictors, exponentially increasing the number of operations required for the oblique RSF versus its axis based counterpart. In addition, standard methods to estimate variable importance (VI) in the RSF are less effective in the oblique RSF, and few methods have been introduced to estimate VI specifically for the oblique RSF.

The aim of this article is to improve the computational efficiency and interpretability of the oblique RSF. In a general benchmark experiment including 31 risk prediction tasks, we show that oblique RSFs with partially trained models have equivalent or superior prediction accuracy and are orders of magnitude more efficient than oblique RSFs with fully trained models in non-leaf nodes. We introduce a method to estimate VI for oblique RSFs and compare its ability to discriminate between signal and noise variables versus standard and state-of-the-art methods. All methods proposed in this article are available in the \texttt{aorsf} R Package.

@@

For each risk prediction task, we completed 10 runs of Monte-Carlo cross validation. In each run, we used a random sample containing 50\% of the available data for training and the remaining 50\% for testing each of the learners described in \secref{sec:learners}. Then, for each learner, we computed the IPA, time-dependent C-statistic, and computational time required to fit a prediction model and compute risk predictions. If any learner failed to obtain predictions on any particular split of data\footnote{For example, when the prediction task was to predict risk of death in the ACTG 320 clinical trial (26 events total), some splits did not leave enough events in the training data to fit complex learners such as the neural network}, the results for that split were omitted from downstream analyses.

@@

After collecting data from 10 replications of Monte-Carlo cross validation for all 14 learners in all 31 risk prediction tasks, we analyzed the resulting 4,340 observations of IPA and, separately, time-dependent C-statistic, using a Bayesian linear mixed model. Our approach follows the ideas described by \citet{benavoli2017time} and \citet{tidymodels}, who developed guidelines on making statistical comparisons between learners using resampling and Bayesian methods. Specifically, we fit two models: $$\text{IPA} = \widehat{\gamma}_0 + \widehat{\gamma} \cdot \text{learner} + (1\,|\, \text{data/run}) $$ and $$\text{C-stat} = \widehat{\gamma}_0 + \widehat{\gamma} \cdot \text{learner} + (1\,|\, \text{data/run}).$$ Random intercepts for specific splits of data (\ie, \texttt{run} in the model formula) were nested within datasets. The intercept, $\widehat{\gamma}_0$, was the expected value using \texttt{aorsf-fast} so that the coefficients in $\widehat{gamma}$ could be interpreted as expected differences between \texttt{aorsf-fast} and other learners. Default priors from \texttt{rstanarm} were applied for model fitting \citep{rstanarm}.

\paragraph{Hypothesis testing} For both the IPA and time-dependent C-statistic, we conducted equivalence and inferiority tests based on a 1 point region of practical equivalence. More specifically, we concluded that two learners had practically equivalent IPA or time-dependent C-statistic if there was a 95\% or higher posterior probability that the absolute difference in the relevant metric was less than 1. We concluded that one learner was weakly superior when there was $\geq$ 0.95 posterior probability that the difference in the relevant metric was non-zero, and concluded superiority when when there was $\geq$ 0.95 posterior probability that the difference in the relevant metric was 1 or more.

\subsubsection{Results}

@@

Compared to learners that were not oblique RSFs, \texttt{aorsf-fast} had the highest IPA in  out of 31 risk prediction tasks, with an overall mean IPA of  (Figure \ref{fig:bm_pred_viz_ibs}). Compared to the learner with the second highest mean IPA (\texttt{rsf-standard}), \texttt{aorsf-fast}'s mean was 1.21 points higher, a relative increase of 9.98\%. The posterior probability of \texttt{aorsf-fast} and \texttt{aorsf-cph} having practically equivalent expected IPA was 0.97, and the posterior probability of \texttt{aorsf-fast} having a superior IPA to other learners ranged from 0.65 (versus \texttt{rsf-standard}) to $>$0.999 (versus several other learners; see Figure \ref{fig:bm_pred_model_viz_ibs})

@@

Compared to learners that were not oblique RSFs, \texttt{aorsf-fast} had the highest time-dependent C-statistic in  out of 31 risk prediction tasks, with an overall mean of  (Figure \ref{fig:bm_pred_viz_cstat}). Compared to the learner with the second highest mean C-statistic (\texttt{obliqueRSF-net}), \texttt{aorsf-fast}'s mean was 0.681 points higher, a relative increase of 0.887\%. The posterior probability of \texttt{aorsf-fast} and \texttt{aorsf-cph} having practically equivalent expected time-dependent C-statistics was 0.99, and the posterior probability of \texttt{aorsf-fast} having a superior time-dependent C-statistic versus other learners ranged from 0.01 (versus \texttt{obliqueRSF-net}) to $>$0.999 (versus several other learners; see Figure \ref{fig:bm_pred_model_viz_cstat})

@@

\subsubsection{Results}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\begin{landscape}
\begin{longtable}[t]{lcclcc}
\caption{\label{tab:unnamed-chunk-4}Data sets used for numeric experiments \label{tab:datasets}}\\
\toprule
Label & N observations & N predictors & Outcome & N Events & \% Censored\\
\midrule
VA lung cancer trial & 137 & 8 & Death & 128 & 6.57\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Recurrence & 468 & 49.6\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash Colon cancer} & \multirow{-2}{*}{\centering\arraybackslash 929} & \multirow{-2}{*}{\centering\arraybackslash 12} & Death & 452 & 51.3\\
\cmidrule{1-6}\pagebreak[0]
Primary biliary cholangitis & 276 & 19 & Death & 111 & 59.8\\
\cmidrule{1-6}\pagebreak[0]
Movies released in 2015-2018 & 551 & 46 & Gross 1M USD & 522 & 5.26\\
\cmidrule{1-6}\pagebreak[0]
GBSG II & 686 & 10 & Recurrence Or Death & 299 & 56.4\\
\cmidrule{1-6}\pagebreak[0]
Systolic Heart Failure & 2,231 & 41 & Death & 726 & 67.5\\
\cmidrule{1-6}\pagebreak[0]
Serum free light chain & 7,874 & 10 & Death & 2,169 & 72.5\\
\cmidrule{1-6}\pagebreak[0]
Non-alcohol fatty liver disease & 17,549 & 24 & Death & 1,364 & 92.2\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Recurrence & 1,518 & 49.1\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash Rotterdam tumor bank} & \multirow{-2}{*}{\centering\arraybackslash 2,982} & \multirow{-2}{*}{\centering\arraybackslash 11} & Death & 1,272 & 57.3\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & AIDS Diagnosis & 96 & 91.7\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash ACTG 320} & \multirow{-2}{*}{\centering\arraybackslash 1,151} & \multirow{-2}{*}{\centering\arraybackslash 12} & Death & 26 & 97.7\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Cardiovascular Death & 110 & 87.7\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash GUIDE-IT} & \multirow{-2}{*}{\centering\arraybackslash 894} & \multirow{-2}{*}{\centering\arraybackslash 59} & Hf Hospitalization & 288 & 67.8\\
\cmidrule{1-6}\pagebreak[0]
Early breast cancer & 614 & 1,692 & Recurrence Or Death & 134 & 78.2\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Cardiovascular Death & 521 & 94.4\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash SPRINT} & \multirow{-2}{*}{\centering\arraybackslash 9,361} & \multirow{-2}{*}{\centering\arraybackslash 174} & Death & 1,644 & 82.4\\
\cmidrule{1-6}\pagebreak[0]
Heart Transplant & 3,787 & 52 & Graft-Loss Or Death & 500 & 86.8\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Death & 76 & 86.0\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash FCL} & \multirow{-2}{*}{\centering\arraybackslash 541} & \multirow{-2}{*}{\centering\arraybackslash 7} & Relapse & 272 & 49.7\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Death & 963 & 30.4\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash Monoclonal gammopathy} & \multirow{-2}{*}{\centering\arraybackslash 1,384} & \multirow{-2}{*}{\centering\arraybackslash 8} & Malignancy & 115 & 91.7\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Heart Failure & 339 & 95.0\\
\cmidrule{4-6}\nopagebreak
 &  &  & Coronary Heart Disease & 439 & 93.5\\
\cmidrule{4-6}\nopagebreak
 &  &  & Stroke & 292 & 95.7\\
\cmidrule{4-6}\nopagebreak
\multirow{-4}{*}{\raggedright\arraybackslash MESA} & \multirow{-4}{*}{\centering\arraybackslash 6,783} & \multirow{-4}{*}{\centering\arraybackslash 48} & Death & 1,297 & 80.9\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Heart Failure & 2,981 & 78.1\\
\cmidrule{4-6}\nopagebreak
 &  &  & Coronary Heart Disease & 2,282 & 83.2\\
\cmidrule{4-6}\nopagebreak
 &  &  & Stroke & 1,323 & 90.3\\
\cmidrule{4-6}\nopagebreak
\multirow{-4}{*}{\raggedright\arraybackslash ARIC} & \multirow{-4}{*}{\centering\arraybackslash 13,623} & \multirow{-4}{*}{\centering\arraybackslash 41} & Death & 6,662 & 51.1\\
\bottomrule
\end{longtable}
\end{landscape}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\begin{longtable}[t]{lcclc}
\caption{\label{tab:unnamed-chunk-5}Index of prediction accuracy, time-dependent concordance statistic, and computational time required to fit and compute predictions for several learning algorithms across 31 risk prediction tasks.}\\
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Performance metric (SD)} & \multicolumn{2}{c}{Computation time, seconds} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
  & Scaled Brier & C-Statistic & Model fitting & Risk prediction\\
\midrule
\endfirsthead
\caption[]{Index of prediction accuracy, time-dependent concordance statistic, and computational time required to fit and compute predictions for several learning algorithms across 31 risk prediction tasks. \textit{(continued)}}\\
\toprule
  & Scaled Brier & C-Statistic & Model fitting & Risk prediction\\
\midrule
\endhead

@@

\hspace{1em}aorsf-fast & 0.133 (0.114) & 0.775 (0.072) & 0.801 & 0.170\\
\hspace{1em}aorsf-cph & 0.132 (0.114) & 0.774 (0.071) & 1.965 & 0.173\\
\hspace{1em}aorsf-net & 0.131 (0.116) & 0.774 (0.071) & 122.683 & 0.166\\
\hspace{1em}rsf-standard & 0.121 (0.117) & 0.761 (0.075) & 2.089 & 0.252\\
\hspace{1em}cif-rotate & 0.120 (0.130) & 0.758 (0.081) & 39.689 & 8.754\\
\hspace{1em}cif-standard & 0.119 (0.100) & 0.768 (0.070) & 4.254 & 15.360\\
\hspace{1em}obliqueRSF-net & 0.118 (0.086) & 0.774 (0.071) & 376.210 & 24.600\\
\hspace{1em}glmnet-cox & 0.116 (0.123) & 0.763 (0.074) & 0.532 & 0.004\\
\hspace{1em}ranger-extratrees & 0.103 (0.088) & 0.764 (0.068) & 2.519 & 2.159\\
\hspace{1em}cif-extension & 0.099 (0.095) & 0.763 (0.072) & 23.746 & 8.318\\
\hspace{1em}aorsf-random & 0.094 (0.084) & 0.743 (0.064) & 2.088 & 0.178\\
\hspace{1em}xgboost-cox & 0.073 (0.102) & 0.757 (0.088) & 5.270 & 0.005\\
\hspace{1em}nn-cox & 0.046 (0.106) & 0.655 (0.132) & 16.511 & 5.207\\
\hspace{1em}xgboost-aft & -4.78 (4.97) & 0.767 (0.076) & 15.673 & 0.007\\

@@

\hspace{1em}aorsf-random & 0.031 (0.015) & 0.739 (0.031) & 0.396 & 0.036\\
\hspace{1em}ranger-extratrees & 0.030 (0.014) & 0.734 (0.033) & 0.448 & 0.146\\
\hspace{1em}cif-standard & 0.029 (0.028) & 0.740 (0.043) & 0.988 & 4.459\\
\hspace{1em}obliqueRSF-net & 0.027 (0.022) & 0.735 (0.039) & 26.729 & 15.372\\
\hspace{1em}aorsf-cph & 0.026 (0.024) & 0.742 (0.040) & 0.496 & 0.036\\
\hspace{1em}aorsf-fast & 0.026 (0.022) & 0.738 (0.040) & 0.161 & 0.036\\
\hspace{1em}cif-extension & 0.023 (0.015) & 0.711 (0.041) & 9.394 & 4.051\\
\hspace{1em}glmnet-cox & 0.022 (0.020) & 0.741 (0.024) & 0.182 & 0.003\\
\hspace{1em}aorsf-net & 0.021 (0.030) & 0.735 (0.038) & 18.601 & 0.037\\
\hspace{1em}rsf-standard & 0.011 (0.030) & 0.722 (0.044) & 0.149 & 0.065\\
\hspace{1em}xgboost-cox & 0.008 (0.033) & 0.743 (0.035) & 4.586 & 0.003\\
\hspace{1em}cif-rotate & 0.006 (0.036) & 0.726 (0.035) & 15.373 & 3.791\\
\hspace{1em}nn-cox & -0.008 (0.016) & 0.558 (0.129) & 13.641 & 0.549\\
\hspace{1em}xgboost-aft & -7.76 (1.18) & 0.733 (0.029) & 13.603 & 0.007\\

@@

\hspace{1em}aorsf-fast & 0.013 (0.022) & 0.847 (0.054) & 0.104 & 0.022\\
\hspace{1em}aorsf-cph & 0.012 (0.021) & 0.833 (0.061) & 0.335 & 0.021\\
\hspace{1em}aorsf-net & 0.010 (0.032) & 0.829 (0.062) & 13.672 & 0.024\\
\hspace{1em}obliqueRSF-net & 0.009 (0.013) & 0.839 (0.052) & 8.926 & 10.688\\

>>>>

The random survival forest (RSF; \citet{ishwaran2008random, hothorn2006unbiased}) is a supervised learning algorithm that can be used for risk prediction \citep{wang2017selective}, which may reduce the burden of disease by guiding strategies for prevention and treatment in a wide range of medical domains \citep{moons2012riskII, moons2012riskI}. Similar to random forests (RFs) for classification and regression \citep{breiman2001random}, The RSF is a large set of de-correlated and randomized decision trees, with each tree contributing to the ensemble's prediction function. Notable characteristics of the RSF include uniform convergence of its ensemble survival prediction function to the true survival function, first shown by \citet{ishwaran2010consistency} and later by \citet{cui2017consistency} under more general conditions. However, \citet{cui2017consistency} noted that the RSF is at a disadvantage when predictors are correlated and some are not relevant to the censored outcome, which is a strong possibility when large medical databases are leveraged for risk prediction.

A potential approach to improve the RSF when predictors are correlated and some are not relevant to the censored outcome is to use oblique decision trees instead of axis based trees. Axis based trees split data using a single predictor, creating decision boundaries that are perpendicular or parallel to axes of the predictor space \citep[see][Chapter~2]{breiman2017classification}. Oblique trees split data using a linear combination of predictors, creating decision boundaries that are neither parallel nor perpendicular to axes of their contributing predictors \citep[see][Chapter~5]{breiman2017classification}. \citet{menze2011oblique} examined prediction accuracy of RFs in the presence of correlated predictors and found that oblique RFs had substantially higher prediction accuracy compared to axis-based RFs. Similarly, \citet{jaeger2019oblique} found that growing RSFs with oblique rather than axis-based survival trees reduced the RSF's concordance error, with improvements ranging from 2.5\% to 24.9\% depending on the data analyzed.

Oblique trees have at least two notable drawbacks compared to axis-based trees. First, finding a locally optimal oblique decision rule may require exponentially more computation than an axis-based rule. If $p$ predictors are potentially used to split $n$ observations, up to $\mathcal{O}(n^p)$ oblique splits can be assessed versus $\mathcal{O}(n \cdot p)$ axis-based splits \citep{heath1993induction, murthy1994system}. Second, estimating variable importance (VI) using permutation (a standard method for RFs) may be less effective in ensembles of oblique trees, as permuting the values of one predictor may not destabilize decisions that are based on many predictors. Although VI is one of the most widely used strategies to interpret random forests \citep{ishwaran2019standard}, few studies have investigated VI for oblique random forests \citep[see][Section~5]{menze2011oblique}, and fewer have investigated VI specifically for the oblique RSF.

This study makes two contributions to oblique RSFs. First, we reduce their computational cost (\ie, accelerate them) with a scalable algorithm to identify linear combinations of coefficients. In a general benchmark experiment including 31 risk prediction tasks, we show that accelerated oblique RSFs are roughly 500 times faster with equivalent or superior prediction accuracy compared to existing routines to fit oblique RSFs. Second, we improve their interpretability with `negation VI', a method to estimate VI that engages with coefficients in linear combinations of predictors instead of permuting predictor values. In simulation studies where VI is estimate using permutation, analysis of variance (ANOVA; see \citet{menze2011oblique}), and approximations of Shapley values, we find that negation VI improves the oblique RSF's ability to discriminate between correlated signal and noise variables. The accelerated oblique RSF and multiple methods to compute VI for oblique RSFs (permutation, ANOVA, and negation) are available in the \texttt{aorsf} R Package.

% However, there remains considerable potential to improve the RSF in settings where training samples are not large enough to guarantee asymptotic properties or do not align with the conditions required for consistency of the RSF.

% Random forests (RFs) are large sets of de-correlated, randomized decision trees \citep{breiman2001random}, which can be axis based or oblique.Although using oblique rather than axis-based trees to grow a RSF may improve its prediction accuracy,

@@

For each risk prediction task, we completed 5 runs of Monte-Carlo cross validation. In each run, we used a random sample containing 50\% of the available data for training and the remaining 50\% for testing each of the learners described in \secref{sec:learners}. Then, for each learner, we computed the IPA, time-dependent C-statistic, and computational time required to fit a prediction model and compute risk predictions. If any learner failed to obtain predictions on any particular split of data\footnote{For example, when the prediction task was to predict risk of death in the ACTG 320 clinical trial (26 events total), some splits did not leave enough events in the training data to fit complex learners such as the neural network}, the results for that split were omitted from downstream analyses.

@@

After collecting data from 5 replications of Monte-Carlo cross validation for the 14 learners in all 31 risk prediction tasks, we analyzed the resulting 2,170 observations of IPA and, separately, time-dependent C-statistic, using a Bayesian linear mixed model. Our approach follows the ideas described by \citet{benavoli2017time} and \citet{tidymodels}, who developed guidelines on making statistical comparisons between learners using Bayesian models. Specifically, we fit two models: $$\text{IPA} = \widehat{\gamma}_0 + \widehat{\gamma} \cdot \text{learner} + (1\,|\, \text{data/run}) $$ and $$\text{C-stat} = \widehat{\gamma}_0 + \widehat{\gamma} \cdot \text{learner} + (1\,|\, \text{data/run}).$$ Random intercepts for specific splits of data (\ie, \texttt{run} in the model formula) were nested within datasets. The intercept, $\widehat{\gamma}_0$, was the expected value of the outcome using \texttt{aorsf-fast}, making the coefficients in $\widehat{\gamma}$ the expected differences between \texttt{aorsf-fast} and other learners. Default priors from \texttt{rstanarm} were applied for model fitting \citep{rstanarm}.

\paragraph{Hypothesis testing} For both the IPA and time-dependent C-statistic, we conducted equivalence and inferiority tests based on a 1 point region of practical equivalence. More specifically, we concluded that two learners had practically equivalent IPA or time-dependent C-statistic if there was a 95\% or higher posterior probability that the absolute difference in the relevant metric was less than 1. We concluded that one learner was weakly superior when there was $\geq$ 0.95\% posterior probability that the difference in the relevant metric was non-zero, and concluded superiority when when there was $\geq$ 0.95\% posterior probability that the difference in the relevant metric was 1 or more.

\subsubsection{Results} \label{sec:results_pred}

@@

Compared to learners that were not oblique RSFs, \texttt{aorsf-fast} had the highest IPA in  out of 31 risk prediction tasks, with an overall mean IPA of  (Figure \ref{fig:bm_pred_viz_ibs}). Compared to the learner with the second highest mean IPA (\texttt{rsf-standard}), \texttt{aorsf-fast}'s mean was 1.12 points higher, a relative increase of 9.30\%. The posterior probability of \texttt{aorsf-fast} and \texttt{aorsf-cph} having practically equivalent expected IPA was 0.95, and the posterior probability of \texttt{aorsf-fast} having a superior IPA to other learners ranged from 0.63 (versus \texttt{rsf-standard}) to $>$0.999 (versus several other learners; see Figure \ref{fig:bm_pred_model_viz_ibs})

@@

Compared to learners that were not oblique RSFs, \texttt{aorsf-fast} had the highest time-dependent C-statistic in  out of 31 risk prediction tasks, with an overall mean of  (Figure \ref{fig:bm_pred_viz_cstat}). Compared to the learner with the second highest mean C-statistic (\texttt{obliqueRSF-net}), \texttt{aorsf-fast}'s mean was 0.612 points higher, a relative increase of 0.797\%. The posterior probability of \texttt{aorsf-fast} and \texttt{aorsf-cph} having practically equivalent expected time-dependent C-statistics was 0.96, and the posterior probability of \texttt{aorsf-fast} having a superior time-dependent C-statistic versus other learners ranged from 0.02 (versus \texttt{obliqueRSF-net}) to $>$0.999 (versus several other learners; see Figure \ref{fig:bm_pred_model_viz_cstat})

@@

\subsubsection{Results} \label{sec:results_vi}

Overall, \texttt{aorsf-negate} had the highest C-statistic in 21 out of 36 VI tasks, with an overall mean of 75.3 (Figure \ref{fig:bm_vi_viz}). Compared to the VI technique with the second highest overall C-statistic (\texttt{aorsf-anova}), \texttt{aorsf-negate}'s mean was 2.34 points higher, a relative increase of 3.21\%. Among each of the four relevant variable classes, \texttt{aorsf-negate} had the highest mean C-statistic, with the greatest advantage of using \texttt{aorsf-negate} occurring among non-linear and combination variables and the smallest advantage for interaction variables.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/bm_vi_viz-1} \caption[Concordance statistic for assigning higher importance to relevant versus irrelevant variables]{Concordance statistic for assigning higher importance to relevant versus irrelevant variables. Text appears in rows where negation importance obtained the highest concordance, showing absolute and percent improvement over the second best technique.}\label{fig:bm_vi_viz}
\end{figure}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\begin{landscape}
\begin{longtable}[t]{lcclcc}
\caption{\label{tab:unnamed-chunk-5}Data sets used for numeric experiments \label{tab:datasets}}\\
\toprule
Label & N observations & N predictors & Outcome & N Events & \% Censored\\
\midrule
VA lung cancer trial & 137 & 8 & Death & 128 & 6.57\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Recurrence & 468 & 49.6\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash Colon cancer} & \multirow{-2}{*}{\centering\arraybackslash 929} & \multirow{-2}{*}{\centering\arraybackslash 12} & Death & 452 & 51.3\\
\cmidrule{1-6}\pagebreak[0]
Primary biliary cholangitis & 276 & 19 & Death & 111 & 59.8\\
\cmidrule{1-6}\pagebreak[0]
Movies released in 2015-2018 & 551 & 46 & Gross 1M USD & 522 & 5.26\\
\cmidrule{1-6}\pagebreak[0]
GBSG II & 686 & 10 & Recurrence Or Death & 299 & 56.4\\
\cmidrule{1-6}\pagebreak[0]
Systolic Heart Failure & 2,231 & 41 & Death & 726 & 67.5\\
\cmidrule{1-6}\pagebreak[0]
Serum free light chain & 7,874 & 10 & Death & 2,169 & 72.5\\
\cmidrule{1-6}\pagebreak[0]
Non-alcohol fatty liver disease & 17,549 & 24 & Death & 1,364 & 92.2\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Recurrence & 1,518 & 49.1\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash Rotterdam tumor bank} & \multirow{-2}{*}{\centering\arraybackslash 2,982} & \multirow{-2}{*}{\centering\arraybackslash 11} & Death & 1,272 & 57.3\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & AIDS Diagnosis & 96 & 91.7\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash ACTG 320} & \multirow{-2}{*}{\centering\arraybackslash 1,151} & \multirow{-2}{*}{\centering\arraybackslash 12} & Death & 26 & 97.7\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Cardiovascular Death & 110 & 87.7\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash GUIDE-IT} & \multirow{-2}{*}{\centering\arraybackslash 894} & \multirow{-2}{*}{\centering\arraybackslash 59} & Hf Hospitalization & 288 & 67.8\\
\cmidrule{1-6}\pagebreak[0]
Early breast cancer & 614 & 1,692 & Recurrence Or Death & 134 & 78.2\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Cardiovascular Death & 521 & 94.4\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash SPRINT} & \multirow{-2}{*}{\centering\arraybackslash 9,361} & \multirow{-2}{*}{\centering\arraybackslash 174} & Death & 1,644 & 82.4\\
\cmidrule{1-6}\pagebreak[0]
Heart Transplant & 3,787 & 52 & Graft-Loss Or Death & 500 & 86.8\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Death & 76 & 86.0\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash FCL} & \multirow{-2}{*}{\centering\arraybackslash 541} & \multirow{-2}{*}{\centering\arraybackslash 7} & Relapse & 272 & 49.7\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Death & 963 & 30.4\\
\cmidrule{4-6}\nopagebreak
\multirow{-2}{*}{\raggedright\arraybackslash Monoclonal gammopathy} & \multirow{-2}{*}{\centering\arraybackslash 1,384} & \multirow{-2}{*}{\centering\arraybackslash 8} & Malignancy & 115 & 91.7\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Heart Failure & 339 & 95.0\\
\cmidrule{4-6}\nopagebreak
 &  &  & Coronary Heart Disease & 439 & 93.5\\
\cmidrule{4-6}\nopagebreak
 &  &  & Stroke & 292 & 95.7\\
\cmidrule{4-6}\nopagebreak
\multirow{-4}{*}{\raggedright\arraybackslash MESA} & \multirow{-4}{*}{\centering\arraybackslash 6,783} & \multirow{-4}{*}{\centering\arraybackslash 48} & Death & 1,297 & 80.9\\
\cmidrule{1-6}\pagebreak[0]
 &  &  & Heart Failure & 2,981 & 78.1\\
\cmidrule{4-6}\nopagebreak
 &  &  & Coronary Heart Disease & 2,282 & 83.2\\
\cmidrule{4-6}\nopagebreak
 &  &  & Stroke & 1,323 & 90.3\\
\cmidrule{4-6}\nopagebreak
\multirow{-4}{*}{\raggedright\arraybackslash ARIC} & \multirow{-4}{*}{\centering\arraybackslash 13,623} & \multirow{-4}{*}{\centering\arraybackslash 41} & Death & 6,662 & 51.1\\
\bottomrule
\end{longtable}
\end{landscape}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\begin{longtable}[t]{lcclc}
\caption{\label{tab:unnamed-chunk-6}Index of prediction accuracy, time-dependent concordance statistic, and computational time required to fit and compute predictions for several learning algorithms across 31 risk prediction tasks.}\\
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Performance metric (SD)} & \multicolumn{2}{c}{Computation time, seconds} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
  & Scaled Brier & C-Statistic & Model fitting & Risk prediction\\
\midrule
\endfirsthead
\caption[]{Index of prediction accuracy, time-dependent concordance statistic, and computational time required to fit and compute predictions for several learning algorithms across 31 risk prediction tasks. \textit{(continued)}}\\
\toprule
  & Scaled Brier & C-Statistic & Model fitting & Risk prediction\\
\midrule
\endhead

@@

\hspace{1em}aorsf-fast & 0.132 (0.114) & 0.774 (0.072) & 0.833 & 0.173\\
\hspace{1em}aorsf-cph & 0.131 (0.115) & 0.772 (0.072) & 2.477 & 0.174\\
\hspace{1em}aorsf-net & 0.130 (0.118) & 0.772 (0.072) & 131.969 & 0.176\\
\hspace{1em}rsf-standard & 0.121 (0.118) & 0.761 (0.075) & 2.070 & 0.291\\
\hspace{1em}cif-standard & 0.120 (0.101) & 0.768 (0.071) & 4.499 & 15.774\\
\hspace{1em}cif-rotate & 0.119 (0.129) & 0.758 (0.080) & 43.469 & 9.418\\
\hspace{1em}obliqueRSF-net & 0.119 (0.088) & 0.773 (0.071) & 388.086 & 24.864\\
\hspace{1em}glmnet-cox & 0.115 (0.123) & 0.761 (0.074) & 0.693 & 0.004\\
\hspace{1em}ranger-extratrees & 0.104 (0.090) & 0.763 (0.068) & 2.147 & 1.592\\
\hspace{1em}cif-extension & 0.099 (0.095) & 0.764 (0.072) & 23.953 & 8.487\\
\hspace{1em}aorsf-random & 0.095 (0.085) & 0.742 (0.065) & 2.288 & 0.174\\
\hspace{1em}xgboost-cox & 0.076 (0.105) & 0.757 (0.088) & 4.399 & 0.005\\
\hspace{1em}nn-cox & 0.045 (0.101) & 0.648 (0.137) & 16.081 & 5.931\\
\hspace{1em}xgboost-aft & -4.80 (5.11) & 0.767 (0.077) & 14.692 & 0.007\\

@@

\hspace{1em}ranger-extratrees & 0.028 (0.018) & 0.737 (0.042) & 0.053 & 0.155\\
\hspace{1em}aorsf-random & 0.024 (0.019) & 0.732 (0.042) & 0.452 & 0.036\\
\hspace{1em}cif-standard & 0.024 (0.039) & 0.741 (0.060) & 1.460 & 4.523\\
\hspace{1em}cif-extension & 0.022 (0.021) & 0.716 (0.055) & 9.093 & 4.176\\
\hspace{1em}obliqueRSF-net & 0.020 (0.030) & 0.733 (0.056) & 26.723 & 16.653\\
\hspace{1em}aorsf-cph & 0.017 (0.032) & 0.736 (0.056) & 0.452 & 0.037\\
\hspace{1em}aorsf-fast & 0.016 (0.027) & 0.733 (0.054) & 0.153 & 0.034\\
\hspace{1em}glmnet-cox & 0.011 (0.023) & 0.735 (0.034) & 0.180 & 0.002\\
\hspace{1em}aorsf-net & 0.010 (0.041) & 0.731 (0.055) & 19.025 & 0.036\\
\hspace{1em}xgboost-cox & 0.006 (0.046) & 0.747 (0.052) & 3.312 & 0.003\\
\hspace{1em}rsf-standard & 0.004 (0.042) & 0.721 (0.058) & 0.466 & 0.061\\
\hspace{1em}nn-cox & 0.001 (0.007) & 0.589 (0.138) & 10.976 & 0.666\\
\hspace{1em}cif-rotate & -0.002 (0.050) & 0.728 (0.051) & 14.852 & 4.031\\
\hspace{1em}xgboost-aft & -7.31 (0.690) & 0.739 (0.040) & 9.771 & 0.006\\

@@

\hspace{1em}cif-extension & 0.004 (0.012) & 0.794 (0.066) & 12.087 & 5.085\\
\hspace{1em}obliqueRSF-net & 0.003 (0.012) & 0.822 (0.064) & 8.856 & 11.019\\
\hspace{1em}aorsf-fast & 0.001 (0.019) & 0.830 (0.062) & 0.094 & 0.020\\
\hspace{1em}cif-standard & 0.000 (0.022) & 0.797 (0.079) & 1.505 & 4.550\\
